Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct local reduction practices

### 2022-08-07 supp/02-2dgm-rawmat.Rmd
<edit index>

>>>>

## Procrustes ANOVA: Allometry

To assess whether shape changes with size (allometry), Procrustes ANOVAs [@RN1749] were run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions [@RN1756]. A residual randomization permutation procedure (RRPP; n = 10,000 permutations) was used for all Procrustes ANOVAs [@RN1655;@RN11775], which has higher statistical power and a greater ability to identify patterns in the data should they be present [@RN1719].

```{r allomgm1, echo=TRUE, out.width = "100%", dpi = 300, warning=FALSE}
# allometry
fit.size <- procD.lm(shape ~ size, 
                     data = gdf,

                     print.progress = FALSE, 
                     iter = 9999)

# allometry
anova(fit.size)
```

### 2022-08-07 supp/index.Rmd
<edit index>

--- 
title: "Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct reduction practices"
author:
    - Robert Z. Selden, Jr.^[Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; and Cultural Heritage Department, Jean Monnet University, zselden@sfasu.edu] and John E. Dockall^[Stantec, Inc.]
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib]
biblio-style: apalike
link-citations: yes
description: "Pilot study of elliptical biface morphology."
---

@@

Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood object from the ancestral Caddo area. Elliptical bifaces from Jowell Farm were manufactured using Edwards chert, and thought to have been acquired by the Caddo through trade and/or exchange with central Texas groups. Bifaces from the Millsap Cache were reportedly manufactured from Kay County flint, providing for an empirical assessment of whether biface morphology differs as a function of raw material. The sample of elliptical bifaces from each site was divided into two size classes; one reflective of initial production/manufacture (large), and the other associated with resharpening and/or retouch (small). Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shape in the small class. They were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to resharpening and/or retouch practices. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This finding suggests a shared mental template, coupled with a size differential. The subsequent analysis of biface morphology by size class demonstrated that size does not differ by raw material in the large class, but does in the small class. This finding supports the argument that elliptical biface morphology diverges through local resharpening and/or retouch practices, where bifaces manufactured using both raw materials enter their uselife at similar shapes and sizes prior to diverging in size. As expected, greater shape diversity occurs in the small class, where the Jowell Farm bifaces were found to be more morphologically diverse than those from the Millsap Cache. Distinct reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces from the ancestral Caddo area of the American Southeast.

```{r figmap, fig.cap="Geological map illustrating the location of the Millsap Cache and Jowell Farm, the level three ecoregions of the Arkansas Valley (blue), Ouachita Mountains (yellow), South Central Plains (orange), and ancestral Caddo area boundary (white) in the American Southeast."}

@@

Prior to landmarking, elliptical bifaces were oriented with the most heavily retouched edge at top right (Figure \@ref(fig:figlm)). The landmarking protocol uses three landmarks; two horizontal tangents (top/bottom), and one placed at the furthest extent of the edge including the heaviest amount of retouch. Landmarks and semilandmarks were applied using the `StereoMorph` package [@RN8973]. Some bifaces include multiple areas of heavy retouch at the top and bottom of the same lateral edge (Figure \@ref(fig:figlm): b, c, d, and e), while others (Figure \@ref(fig:figlm): f) include one retouched edge at top right and another at bottom left.

```{r figlm, fig.cap="Selected elliptical bifaces from the Jowell Farm site highlighting the most heavily retouched area identified using a modified version of the approach developed and advanced by Miller [-@RN11459], Smallwood [-@RN8310], and Miller and Smallwood [-@RN11460]."}

@@

To identify which edge was most heavily retouched, we employed a modified approach to the flaking index initially developed by Miller [-@RN11459], and later advanced and refined by Smallwood [-@RN8310], then Miller and Smallwood [-@RN11460]. The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length (inclusive of curvature - not a linear metric). The number of flake scars was then divided by the length of the worked edge, with the heaviest worked area identified by the greater value.

The pattern that emerged suggested that the longest reworked edge on the elliptical bifaces was always more heavily retouched. This finding was used to incorporate the images of elliptical bifaces from the Millsap Cache [@RN11461], where the longest reworked edges were oriented at top right prior to landmarking.

@@

To assess whether shape and size differed by `raw material` and `size class`, Procrustes ANOVAs [@RN1749] were run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions [@RN1756]. A residual randomization permutation procedure (RRPP; n = 10,000 permutations) was used for all Procrustes ANOVAs [@RN1655;@RN11775], which has higher statistical power and a greater ability to identify patterns in the data should they be present [@RN1719].

@@

  -   Bifaces _do not differ_ in shape, but **differ significantly** in size by raw material/biogeography.
  -   Bifaces in the _large size class_ recovered from the Millsap Cache and Jowell Farm do not differ in shape or size. This suggests consistent morphology in bifaces from the _large size class_, in support of production-based standardization.
  -   Bifaces in the _small size class_ recovered from the Millsap Cache and Jowell Farm **differ significantly** in size. Thus, while shape remains consistent, size differs, suggesting---at minimum---locally divergent approaches to retouch and/or resharpening.
  -   Morphological disparity in the shape of the _small size class_ is **significantly greater** at Jowell Farm, providing evidence for a more standardized approach to resharpening and/or retouch at Millsap Cache. This finding also provides support for the hypothesis that greater shape diversity is associated with retouch and/or resharpening for elliptical bifaces, rather than production/manufacture.

@@

The shape of elliptical bifaces does not differ significantly in either size class; however, size does. Results of the Procrustes ANOVA support the hypothesis for divergent morphology. The analysis of morphological disparity yielded support for the interpretation that greater morphological diversity is associated with resharpening and/or retouch, while providing additional support for morphological consistency in production/manufacture.

@@

Our thanks to the Caddo Nation of Oklahoma, the Caddo Nation Tribal Council, Tribal Chairman, and Tribal Historic Preservation Office for permission and access to NAGPRA and previously repatriated collections. Our gratitude is also extended to Marybeth Tomka and Lauren Bussiere at the Texas Archeological Research Laboratory for their assistance with access to the bifaces and the associated records, and to Scott Hammerstedt and Debra K. Green at the Oklahoma Archeological Survey for their assistance with records requests. Thanks also to John Harman for access to the DStretch plugin for ImageJ that was useful in the analysis of flake scars, and to Harry J. Shafer, Hiram F. (Pete) Gregory, Christian S. Hoggard, and David K. Thulman for their comments and constructive criticisms on the ongoing analyses of Caddo biface morphology, and to Emma Sherratt, Kersten Bergstrom, Dean C. Adams, and Michael L. Collyer for their constructive criticisms, general comments, and suggestions throughout the development of this research program.

>>>>

--- 
title: "Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct local reduction practices"
author:
    - Robert Z. Selden, Jr.^[Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; and Cultural Heritage Department, Jean Monnet University, zselden@sfasu.edu] and John E. Dockall^[Stantec, Inc.]
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib]
biblio-style: apalike
link-citations: yes
description: "Pilot study of elliptical biface morphology."
---

@@

Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood lithic artefact from the ancestral Caddo area. Elliptical bifaces from Jowell Farm were manufactured using Edwards chert, and are thought to have been acquired by the Caddo through trade or exchange with central Texas groups. Bifaces from Millsap Cache were reportedly manufactured from Kay County flint, providing for an empirical test of whether biface morphology differs as a function of raw material. The sample of elliptical bifaces from each site was also divided into two size classes; one reflective of initial production (large), and the other associated with retouch (small). Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shape in the small class. They were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This finding suggests a shared mental template, coupled with a size differential. The subsequent analysis of biface morphology by size class demonstrated that size does not differ by raw material in the large class, but does in the small class. This finding supports the argument that elliptical biface morphology diverges through local retouch practices. As expected, greater shape diversity occurs in the small class, where Jowell Farm bifaces were found to be more morphologically diverse than those from Millsap Cache. Distinct local reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces from the ancestral Caddo area of the American Southeast.

### Keywords {-}

American Southeast, Caddo, Jowell biface, archaeology, museum studies, digital humanities, geometric morphometrics

```{r figmap, fig.cap="Geological map illustrating the location of the Millsap Cache and Jowell Farm, and the level three ecoregions of the Arkansas Valley (blue), Ouachita Mountains (yellow), and South Central Plains (orange) in the ancestral Caddo area (white) of the American Southeast."}

@@

Prior to landmarking, elliptical bifaces were oriented with the most heavily retouched edge at top right (Figure \@ref(fig:figlm)). The landmarking protocol uses three landmarks; two horizontal tangents (top/bottom), and the third placed at the furthest extent of the edge that includes the heaviest amount of retouch. Landmarks and semilandmarks were applied using the `StereoMorph` package [@RN8973]. Some bifaces include multiple areas of heavy retouch at the top and bottom of the same lateral edge (Figure \@ref(fig:figlm): b, c, d, and e), while others (Figure \@ref(fig:figlm): f) include one retouched edge at top right and another at bottom left.

```{r figlm, fig.cap="Selected elliptical bifaces from the Jowell Farm site highlighting the most heavily retouched area identified using a modified version of the flaking index developed and advanced by Miller [-@RN11459], Smallwood [-@RN8310], and Miller and Smallwood [-@RN11460]."}

@@

To identify which edge was most heavily retouched, we employed a modified approach to the flaking index initially developed by Miller [-@RN11459], and later advanced and refined by Smallwood [-@RN8310], then Miller and Smallwood [-@RN11460]. The modified approach uses counts of flake scars from each edge in the two most heavily worked areas, paired with a measure of edge length (inclusive of curvature - this is not a linear metric). The number of flake scars was then divided by the length of the worked edge, with the heaviest worked edge identified by the greater value.

The longest reworked edge on the elliptical bifaces was almost always more heavily retouched. This finding was used to incorporate the images of elliptical bifaces from Millsap Cache [@RN11461], where the longest reworked edge was oriented at top right prior to landmarking.

@@

To assess whether `shape` changes with `size` (allometry), and whether shape and size differed by `raw material` and `size class`, Procrustes ANOVAs [@RN1749] were run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions [@RN1756]. A residual randomization permutation procedure (RRPP; n = 10,000 permutations) was used for all Procrustes ANOVAs [@RN1655;@RN11775], which has higher statistical power and a greater ability to identify patterns in the data should they be present [@RN1719].

@@

  -   Bifaces _do not differ_ in shape, but **differ significantly** in size by raw material/biogeography.
  -   Allometry (shape change with size) was not significant.
  -   Bifaces in the _large size class_ recovered from the Millsap Cache and Jowell Farm do not differ in shape or size. This suggests consistent morphology in bifaces from the _large size class_, in support of production-based standardization.
  -   Bifaces in the _small size class_ recovered from the Millsap Cache and Jowell Farm **differ significantly** in size. Thus, while shape remains consistent, size differs, suggesting---at minimum---locally divergent approaches to retouch and/or resharpening.
  -   Morphological disparity in the shape of the _small size class_ is **significantly greater** at Jowell Farm, providing evidence for a more standardized approach to resharpening and/or retouch at Millsap Cache. This finding also provides support for the hypothesis that greater shape diversity is associated with retouch and/or resharpening for elliptical bifaces, rather than production/manufacture.

@@

The shape of elliptical bifaces does not differ significantly in either size class; however, size does. Results of the Procrustes ANOVAs support the hypothesis for divergent morphology. The analysis of morphological disparity yielded support for the interpretation that greater morphological diversity is associated with local retouch, while providing additional support for morphological consistency related to production.

@@

Our thanks to the Caddo Nation of Oklahoma, the Caddo Nation Tribal Council, Tribal Chairman, and Tribal Historic Preservation Office for permission and access to NAGPRA and previously repatriated collections. Our gratitude is also extended to Marybeth Tomka and Lauren Bussiere at the Texas Archeological Research Laboratory for their assistance with access to the bifaces and the associated records, to Sergio Ayala and the Gault School of Archaeological Research for access to the UV light, and to Scott Hammerstedt and Debra K. Green at the Oklahoma Archeological Survey for their assistance with records requests. Thanks also to John Harman for access to the DStretch plugin for ImageJ that was useful in the analysis of flake scars, and to Harry J. Shafer, Hiram F. (Pete) Gregory, Christian S. Hoggard, and David K. Thulman for their comments and constructive criticisms on the ongoing analyses of Caddo biface morphology, as well as Emma Sherratt, Kersten Bergstrom, Dean C. Adams, and Michael L. Collyer for their constructive criticisms, general comments, and suggestions throughout the development of this research program.
